---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Analysis of DNA and cellular microbiome community standards

## Overview

An important aspect of microbiome profiling studies is to assess the accuaracy of the profiling work-flow. Of course, biases can be introduced at multiple levels of analysis includingn sample collection, DNA extraction, library preparation and sequencing. While we may not be able to mitigate all of the biases, we are able to assess their impact by processing and sequencing a set of microbiome standards. We currently utilise the zymoBIOMICS community standard that consists of 8 bacterial strains and 2 fungal strains. We are able to assess biases at two levels:

* Efficiency of DNA extraction using the cellular standards
* Effciency and accuracy of sequencing and informatics workflows (somewhat confounded) using DNA standards.

Information on the standards can be found [here](https://files.zymoresearch.com/protocols/_d6305_d6306_zymobiomics_microbial_community_dna_standard.pdf).


## Data analysis

The analysis below contains two methods that we use (and are recommended) to analyse the community standards. The first is an analysis of the true postive rate. This determines whether we pull back everything we expect in addition to any false positives. The second method is looking at how accurately the composition (i.e. the % of reads assigned to each standard taxon) represents the expected composition of the community standard. Below is a table of the taxa that are present in the community standard and the expected compostion based on either 16S rRNA amplicon sequencing analysis or shotgun metagenomics analysis.


|Species                   | 16S rRNA abundance % | Shotgun metagenomics abundance % |
|--------------------------|----------------------|--------------------------------- |
|Pseudomonas aeruginosa    | 4.2                  | 6.1                              |
|Escherichia coli          | 10.1                 | 8.5                              |
|Salmonella enterica       | 10.4                 | 8.7                              |
|Lactobacillus fermentum   | 18.4                 | 21.6                             |
|Enterococcus faecalis     | 9.9                  | 14.6                             |
|Staphylococcus aureus     | 15.5                 | 15.2                             |
|Listeria monocytogenes    | 14.1                 | 13.9                             |
|Bacillus subtilis         | 17.4                 | 10.3                             |
|Saccharomyces cerevisiae  | NA                   | 0.57                             |
|Cryptococcus neoformans   | NA                   | 0.37                             |


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(out.extra = '')
knitr::opts_knit$set(root.dir=".")
#knitr::opts_chunk$set(fig.pos = 'H')
```

```{r load.libraries, echo=FALSE, message=FALSE}
library(ggplot2)
library(dplyr)
library(gridExtra)
library(RColorBrewer)
library(reshape)

# configuration
source("OCMS_zymoBIOMICS_config.R")
```

```{r annotations, echo=FALSE, message=FALSE}

# build annotations and abundances depending on the methods and
# databases that were used in the analysis

#####################################################################
# NCBI data for 16S analysis
#####################################################################

anno_ncbi_16s <- data.frame(species = c("Pseudomonas_aeruginosa",
"Escherichia_coli",
"Salmonella_enterica",
"Lactobacillus_fermentum",
"Enterococcus_faecalis",
"Staphylococcus_aureus",
"Listeria_monocytogenes",
"Bacillus_subtilis"),
genus = c("Pseudomonas",
"Escherichia/Shigella",
"Salmonella",
"Lactobacillus",
"Enterococcus",
"Staphylococcus",
"Listeria",
"Bacillus"),
Expected = c(4.2,
10.1,
10.4,
18.4,
9.9,
15.5,
14.1,
17.4))


#####################################################################
# GTDB data for 16S analysis
#####################################################################

anno_gtdb_16s <- data.frame(species = c("Pseudomonas_aeruginosa",
"Escherichia_flexneri",
"Salmonella_enterica",
"Lactobacillus_H_fermentum",
"Enterococcus_faecalis",
"Staphylococcus_aureus",
"Listeria_monocytogenes_B",
"Bacillus_marinus"),
genus = c("Pseudomonas",
"Escherichia",
"Salmonella",
"Lactobacillus_H",
"Enterococcus",
"Staphylococcus",
"Listeria",
"Bacillus"))
anno_gtdb_16s$Expected <- anno_ncbi_16s$Expected

#####################################################################
# NCBI data for shotgun
#####################################################################

anno_ncbi_shotgun <- anno_ncbi_16s
anno_ncbi_shotgun$Expected <- c(6.1, 8.5, 8.7, 21.6, 14.6, 15.2, 13.9, 10.3)
# add fungi
fungi <- data.frame(species = c("Saccharomyces cerevisiae", "Cryptococcus neoformans"),
genus = c("Saccharomyces", "Cryptococcus"),
Expected = c(0.57, 0.37))
anno_ncbi_shotgun <- bind_rows(anno_ncbi_shotgun, fungi)

#####################################################################
# GTDB data for shotgun
#####################################################################

# NB - The fungi aren't present in the GTDB so have to
# sort the abundances accordingly
anno_gtdb_shotgun <- anno_gtdb_16s
gtdb_abundance <- anno_ncbi_shotgun$Expected + 0.1175
anno_gtdb_shotgun$Expected <- gtdb_abundance[1:8]
```

```{r utility_functions, echo=FALSE, message=FALSE}

#####################################################################
#####################################################################
#####################################################################

relab <- function(counts){

    relab <- (sweep(counts, 2, colSums(counts), "/"))*100
    return(relab)
    }

#####################################################################
#####################################################################
#####################################################################

truePosRate <- function(relab, annotations, level="species", cutoff=0.01){

    # sort out rownames to match annotations
    if (level == "species"){
        shortnames <- unlist(strsplit(rownames(relab), "\\|"))
        shortnames <- gsub("s__", "", shortnames[seq(7, length(shortnames), 7)])
        
        # we can't use unassigned
        relab <- relab[shortnames != "unassigned",]
        shortnames <- shortnames[shortnames != "unassigned"]
        
        rownames(relab) <- shortnames
    }else if (level == "genus"){
        shortnames <- unlist(strsplit(rownames(relab), "\\|"))
        shortnames <- gsub("g__", "", shortnames[seq(6, length(shortnames), 7)])
        relab <- aggregate(relab, list(shortnames), sum)
        rownames(relab) <- relab$Group.1
        relab <- relab[,2:ncol(relab)]
    }

    # calculate true positive rate
    result <- list()
    for (i in 1:ncol(relab)){
    ab <- data.frame(taxon=rownames(relab), abundance = relab[,i])
    ab <- ab[ab$abundance >= cutoff,]
    ab <- ab[order(ab$abundance, decreasing=TRUE),]

    trues <- 0
    falses <- 0
    tps <- c()
    labels <- c()
    for (j in 1:nrow(ab)){
        if (ab[j,]$taxon %in% annotations[,level]){
            trues = trues + 1
            labels <- append(labels, "TP")
        }
        else{
            falses = falses + 1
            labels <- append(labels, "FP")
        }
        tp <- trues/(trues + falses)
        tps <- append(tps, tp)
    }
    sample <- rep(colnames(relab)[i], length(tps))
    res <- data.frame(rank = seq(1:length(tps)),
    true.pos.rate = tps,
    sample = sample,
    label = labels
    )
    result[[i]] <- res
    }
    result <- bind_rows(result)
}

```

```{r choose.annotations, echo=FALSE, message=FALSE}

if (annotations == "NCBI" & sequence_type == "16S"){
annotations = anno_ncbi_16s}else if (annotations == "GTDB" & sequence_type == "16S"){
annotations = anno_gtdb_16s}else if (annotations == "NCBI" & sequence_type == "shotgun"){
annotations = anno_ncbi_shotgun}else if (annotations == "GTDB" & sequence_type == "shotgun"){
annotations = anno_gtdb_shotgun}

```

```{r read.data, echo=FALSE, message=FALSE}

counts <- read.csv(counts_table, header=TRUE, stringsAsFactors=FALSE, sep="\t", row.names=1)

# subset based on column names
counts <- counts[,standards_columns]
relabundance <- relab(counts)
```

# True positive rate

Below is a plot of the true positive rate (ordered by abundance high -> low).  

## Species

```{r tp.rate.se, echo=FALSE, message=FALSE, fig.height=5, fig.width=10}

tp.rate.sp <- truePosRate(relabundance, annotations, level="species", cutoff=species_cutoff)
ggplot(tp.rate.sp, aes(x=rank, y=true.pos.rate, colour=label, group=sample)) + geom_point() + theme_bw() + ylab("TP / (TP + FP)") + scale_colour_manual(values=c("grey", "purple")) + facet_wrap(~sample, scale="free")
```


## Genus

```{r tp.rate.ge, echo=FALSE, message=FALSE, fig.height=5, fig.width=5}
tp.rate.ge <- truePosRate(relabundance, annotations, level="genus", cutoff=genus_cutoff)
ggplot(tp.rate.ge, aes(x=rank, y=true.pos.rate, colour=label, group=sample)) + geom_point() + theme_bw() + ylab("TP / (TP + FP)") + scale_colour_manual(values=c("grey", "purple")) + facet_wrap(~sample, scale="free")
```


# Composition analysis

It is useful to visualise how the predicted composition of the standards compares with the expected composition. Below we plot this at species and genus levels.

## Species

```{r composition.species, echo=FALSE, message=FALSE, fig.height=5, fig.width=7}

toplot.sp <- relabundance
species <- unlist(strsplit(rownames(toplot.sp), "\\|"))
species <- gsub("s__", "", species[seq(7, length(species),7)])
toplot.sp$species <- species
known.sp <- merge(annotations, toplot.sp, by.x="species", by.y="species", all.x=TRUE, all.y=FALSE)

# set up colours
colours.sp <- brewer.pal(n=nrow(known.sp), name = "Set3")
known.sp.m <- melt(known.sp)
ggplot(known.sp.m, aes(x=variable, y=value, fill=species)) + geom_bar(stat="identity") + theme_bw() + scale_fill_manual(values=colours.sp) + xlab("") + ylab("% reads")
```

## Genus

```{r composition.genus, echo=FALSE, message=FALSE, fig.height=5, fig.width=7}

toplot.ge <- relabundance
genus <- unlist(strsplit(rownames(toplot.sp), "\\|"))
genus <- gsub("g__", "", genus[seq(6, length(genus),7)])
toplot.ge <- aggregate(toplot.ge, list(genus), sum)
rownames(toplot.ge) <- toplot.ge$Group.1
toplot.ge <- toplot.ge[,2:ncol(toplot.ge)]
toplot.ge$genus <- rownames(toplot.ge)
known.ge <- merge(annotations, toplot.ge, by.x="genus", by.y="genus", all.x=TRUE, all.y=FALSE)

# set up colours
colours.ge <- brewer.pal(n=nrow(known.ge), name = "Set3")
known.ge.m <- melt(known.ge)
ggplot(known.ge.m, aes(x=variable, y=value, fill=species)) + geom_bar(stat="identity") + theme_bw() + scale_fill_manual(values=colours.ge) + xlab("") + ylab("% reads")

```

# Correlation analysis

In addition to visualising the composition it is interesting to look at the correlations between taxa abundances and expected taxa abundances.

## Species

```{r correlation.species, echo=FALSE, message=FALSE, fig.height=5, fig.width=10}
grobs.sp <- list()
for (i in 4:ncol(known.sp)){
sample <- colnames(known.sp)[i]
dat <- data.frame(Expected = known.sp$Expected)
dat[,sample] <- unlist(known.sp[,i])
p <- ggplot(dat, aes_string(x=sample, y="Expected")) + geom_point() + theme_bw() + stat_smooth(method="lm")
grobs.sp[[i-3]] <- p
}
grid.arrange(grobs=grobs.sp, ncol=2)

```


## Genus

```{r correlation.genus, echo=FALSE, message=FALSE, fig.height=5, fig.width=10}
grobs.ge <- list()
for (i in 4:ncol(known.ge)){
sample <- colnames(known.ge)[i]
dat <- data.frame(Expected = known.ge$Expected)
dat[,sample] <- unlist(known.ge[,i])
p <- ggplot(dat, aes_string(x=sample, y="Expected")) + geom_point() + theme_bw() + stat_smooth(method="lm")
grobs.ge[[i-3]] <- p
}
grid.arrange(grobs=grobs.ge, ncol=2)

```
